Hybrid Threads for the Parrot Virtual Machine

### 2012-06-23 Paper.tex
Remove target page count from headlines
First round of fixes suggested by H. Dobler++

\chapter{Thanks (1)}

Andrew "whiteknight" Whitworth for laying the foundations for my work, being of tremendous help at all stages of the project and for fixing some of my bugs.

Nat "Chandon" Tuck for implementing green thread basics on which my work is based on.

Christoph "cotto" Otto for bringing me into the Parrot project.

Brian "benabik" Gernhardt for lurking in the IRC channel and at least trying to help with my many questions.

Markus "zimmski" Zimmerman for correcting many errors.

\chapter{Abstract (1)}

Parrot is a runtime for dynamically typed programming languages. Despite several attempts in it's 10 years of history, it does provide any support for multithreaded computation.

This paper shows a way to implement threading support in the runtime using a hybrid between lightweight (``green'') threads and operating system threads. These lightweight threads are used as messages in a system where reading shared data is allowed but only the one owner thread may write to it. The implementation is described in detail and benchmarks presented demonstrating the viability of this design.

\begin{german}
\chapter{Kurzfassung (1)}

Parrot ist eine Laufzeitumgebung f{\"u}r dynamisch typisierte Programmiersprachen. Trotz mehrerer Anl{\"a}ufe in seiner zehnj{\"a}hrigen Geschichte, bietet es keine Unterst{\"u}tzung f{\"u}r Multithreading.

Diese Arbeit zeigt einen Weg um Threadingunterst{\"u}tzung in der Laufzeitumgebung zu implementieren mit Hilfe eines hybriden Systems aus leichtgewichtigen und Betriebssystem- Threads. Diese leichtgewichtigen Threads werden als Nachrichten in einem System benutzt, in dem das Lesen von gemeinsamen Daten erlaubt ist, aber nur der eine besitzende Thread darauf schreiben darf. Die Implementierung wird detailiert beschrieben und mit Benchmarks die Brauchbarkeit des Designs gezeigt.

@@

\chapter{Introduction and Motivation (3)}

\begin{wrapfigure}{r}{0.5\textwidth}
A simple example of Perl:
\include{perlexample}
\end{wrapfigure}
%
On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a very flexible and widely used programming language but had started to show its age and suffered from early design decisions. The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility\cite{Masak40451}. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions\cite{Developer}. Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot\cite{Perl}.

Parrot evolved from being just the interpreter for the new version of Perl to being a language independent virtual machine providing features like garbage collection, exception handling and dynamic typing. At the time when the Parrot project started, the Java and .NET virtual machines were widely used, but both targeted statically typed languages. Parrot thus filled a quickly growing niche.

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support\cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitly synchronized. This was similar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitly shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

Being born at a time when Perl 6 still looked much more similar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model. Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like Software Transactional Memory failed. For example Parrot has never supported running multiple threads and having garbage collection at the same time.

@@

In the year 2005 development of faster CPUs (Central Processing Units) shifted from increased speed of a single core to adding more cores. Modern processors contain up to 12 cores with even mobile phones having up to four. To utilize a modern CPU's power, code needs to be run in parallel. In UNIX (and thus Perl) tradition, this is accomplished using multiple processes being a good solution for many use cases. For many others like Perl 6's auto threading of hyper operators, the cost of process setup and communication would be prohibitively high except for very large data sets.

@@

Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption entirely to the user. A high level language like Perl 6 on the other hand offers complex and compound data types, handles garbage collection and a very dynamic object system. Even seemingly simple things like a method call can become very complex sequences. In a statically typed programming language the definition of a class is immutable. Thus calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it. Calling the same method again can then even omit the first two steps since their results cannot change.

In a dynamic language, the object may change its class at runtime. The inheritance hierarchy of the class may be changed by adding or removing parent classes. Methods may be added to or removed from classes (or objects) at runtime and even the way to find a method of a class may change. So a simple method call results in the following steps:
%
\begin{itemize}
\item determining the class of the object
\item determining the method resolution method of the class
\item finding the actual method to call
\item calling the method
\end{itemize}
%
These steps have to be repeated for every following method call, because their results may change any time. In a threaded environment, a thread running in parallel may change the underlying data and meta data in between those sequences and even between those steps. As a consequence, this meta data has to be protected from corruption introducing the need for many locks in a very performance critical area.

Many interpreters for dynamic languages like Python or Ruby handle this problem by using a global interpreter lock to effectively serialize all operations. This is a proven and reliable way but again leaves much of the hardware's potential unused.

@@

During these years of back and forth and failed attempts of adding threading support to Parrot, the Perl 6 specification evolved to a point where the largest parts of the language were covered and its features implemented in the compilers. The lack of concurrency primitives in Parrot however prevents any progress in the area of concurrency support.

Before the work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.

This paper suggests a new approach based on a hybrid threading system. So called green threads are used to simplify the implementation of a nearly lock free multithreading implementation. This approach is based on a design by Andrew Whitworth and Nat Tuck. The goal of this paper is to demonstrate the advantages of this model and produce a working implementation which can be used to investigate the performance characteristics of a hybrid threading system.

\chapter{Concurrency in other Programming Platforms (5)}

This chapter is talking about programming platforms. A platform is seen as a combination of a programming language and a runtime. For example for the Python programming language there are multiple runtimes with different implementations of threading support.

@@

In Java responsibility for preventing concurrency issues is obliged on the user code. The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself. The class library provides the user with high level data structures explicitly designed for multithreaded scenarios.

Java version 1.1 used ``green threads'' to support multithreaded execution of Java programs. Green threads are threads simulated by the virtual machine but unable to use more than one CPU core for processing. Details are described in chapter~\ref{cha:green_threads}. Version 1.2 introduced native OS (Operating System) thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.

@@

The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitly type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.

To actually run Python code in parallel, multiple processes have to be used. The \textit{multiprocessing} module provides support for spawning processes using an API similar to the threading module\cite{PythonMultiProcessingManual}. Since processes may not directly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.

\chapter{Parrot (4)}

Parrot consists of the virtual machine (VM, also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit). This paper concentrates on the virtual machine itself. The interpreter itself is written in C. Example code and test cases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.

Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based. A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top. Thus the operands are chosen implicitly by ordering of the operations allowing the opcodes to be quite small. In a register based VM on the other hand each operation has to specify the operands explicitly. Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}. The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.

The current design uses four sets of registers with each and unlimited number of usable registers. The four sets are:
%
\begin{itemize}
\item integer
\item floating point
\item string
\item polymorphic container (PMC)
\end{itemize}

The first three should be self explanatory. Strings in Parrot are a low level type with the interpreter handling all memory allocation issues and Unicode encoding. String values themselves are immutable.

\section{PMCs}
\label{sec:PMCs}

PMCs (PolyMorphic Container) are containers for all higher level types such as objects, arrays, hash tables or code. Thus they are very similar to Python's PythonObject types. PMCs are defined by C structs and are fully garbage collected. Their definition looks like:
\begin{CCode}
struct PMC {
    Parrot_UInt    flags;
    VTABLE         *vtable;             /* Pointer to vtable. */
    DPOINTER       *data;               /* Pointer to attribute structure. */
    PMC            *_metadata;          /* Pointer to metadata PMC. */
};
\end{CCode}
In short they contain a pointer to a type specific data structure and some meta data. The vtable is where the type's behavior is defined. It contains a long list of function pointers forming Parrot's unified type interface. This interface is a union of numeric, string, array, hash and object like behaviors. For example the  \textit{get\_integer} function returns an integer value for the data type. For a simple \textit{int} it is the contained value. For an array it may be the number of elements. The \textit{find\_method} function makes user defined methods of objects possible.

In this way, a language implementer can define the basic types of the language and available operations on them. For each vtable entry, there is a corresponding op code in Parrot's bytecode. Thus an \textit{inc \$P0} instruction is calling the \textit{increment} vtable function of the PMC contained in the \textit{\$P0} register.

PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compilation.

\section{ParrotInterpreter}

The interpreter itself is represented by a C struct called \textit{parrot\_interp\_t}. This structure contains pointers to the garbage collector's runtime data, the loaded types, vtables, the runloop, the current continuation and other global data. A pointer to this structure is passed to almost every function as the first parameter.

@@

Control flow in Parrot is modeled using continuation passing. A continuation is a data structure containing the state of a program at a given point in its execution. It contains all information needed to continue a program in a certain state, e.g. a call stack, the instruction pointer and contents of local variables. ``Calling'' a continuation means restoring the state encapsulated in the continuation.

When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of its parameters. Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them. Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and its value returned unchanged, the outer function can just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as described in chapter~\ref{cha:green_threads} is a very important piece in implementing green threads.

@@

A runloop is in principle the inner most loop executing bytecode in Parrot. It is also a data structure containing some information needed to support exception handling. When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started. Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code calling a previously defined exception handler in user code.

@@

Before entering a runloop, the \textit{setjmp} C function is used to save the current stack position and register contents to a data structure stored in the runloop meta data. This is effectively creating something resembling a continuation at C level. When an exception is created within the interpreter, the runloop stack is searched for the runloop containing a suitable exception handler. \textit{longjmp} is then used to unwind the call stack up to the point where setjmp was called and the call environment restored. While having great similarities with continuations, this mechanism is more limited. It only allows to jump back to a point higher in the call stack.

\section{Garbage Collector}

Parrot supports different GC (Garbage Collector) implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
\begin{itemize}
\item Inf: a GC for debugging purposes never collecting any garbage
\item MS: a basic mark and sweep implementation
\item MS2: a non-recursive mark and sweep implementation
\item GMS: a generational, non-compacting, mark and sweep GC
\end{itemize}
with GMS being the default.

A mark and sweep garbage collector operates in two phases. In the mark phase, starting from a known root set of objects, the GC follows pointers down the object tree, marking each encountered object as alive. In the sweep phase, all objects not being alive are destroyed and the live flags reset.

In a process with many objects, having to traverse the whole tree may take a considerable amount of time. To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object has lived, the lower the chances are that it will become unused. So the objects are partitioned into different generations. The youngest generation would always be traversed while the older generations would be handled much less frequently or even never more at all.

\section{Historical development (1)}

The remainder of the previous threading implementation has been removed with the merging of the kill\_threads branch on September, 21st 2011. Previously much threading related code has been removed to clean up the code and improve performance. Since the existing threading support was known to be unreliable and seriously flawed, this was no trade off.

A year before, Nat Tuck began working on a green\_threads branch during his Google Summer of Code internship. The feature got prototyped using pure PIR and then implemented in Parrot's core. He got it to work in simple cases and started to work on operating system thread support but the internship ended before the code was ready to be merged into the master branch. The code lay dormant until the work on this paper started about a year later.

\chapter{Green threads (3)}
\label{cha:green_threads}

@@

Operating system threads are like multiple processes running at the same time but share their memory. The operating system is responsible for giving each of these threads CPU time to run. Operating system threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.

@@

"Green threads" or "lightweight threads" are threads managed by the virtual machine instead of by the operating system. For better distinction green threads are henceforth referred to as "tasks". These tasks only run pseudo parallel. The virtual machine contains a scheduler and support for preempting running tasks.
Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
This is an implementation of the many-to-one threading model very similar to what an operating system running on a system with a single CPU core does.
%
Advantages of green threads are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item Green threads are very light weight with low memory overhead and close to zero creation time.
\item They do not depend on OS threading support
\item The interpreter controls the point of preemption of a green thread.
\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
\item Garbage collection can be implemented like in a single threaded process.
\item Critical sections can simply be protected by disabling the scheduler until the code is clear of the section.
\end{itemize}
%
Disadvantages include:
%
\begin{itemize}
\item They do not allow more than one CPU core to be used for computations.
\item Blocking calls like I/O block the whole interpreter including other green threads.
\item The interpreter has to contain logic and timers to control green threads.
\end{itemize}

@@

\chapter{Design of Hybrid Threads (3)}
\label{cha:design}

@@

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread. Thus the words ``thread'' and ``interpreter'' are used interchangeably as there is a 1:1 relationship between them. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

As described in the introduction, cross thread writes to shared data may endanger the internal consistency of the interpreter. Traditionally the response to this problem is the use of locks of varying granularity. Fine grained locking allows code to run well parallelized but taking and releasing locks also costs performance. It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate. Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment\cite{LockingInOSKernels}. While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.

@@

The solution for these problems implemented for this paper is to side step them altogether by disallowing write access to shared data. The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started. The interpreter would then create proxy objects for these variables which the task can use to access the data. These proxies contain references to the original objects. They use these references to forward all reading vtable functions to the originals. Write access on the other hand would lead to a runtime error.

@@

For threads to be able to communicate with their creators and other threads, they still need to write to shared data. This is where green threads come into play. Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable. This task would be scheduled on the interpreter owning this variable. To reduce latency, the task is flagged to run immediately. The data owning interpreter would preempt the currently running task and process the new write task. Put another way, the data owning interpreter is simply told what to write to its variables, so other threads don't have to.

@@

\chapter{Implementation of Hybrid Threads (13)}

\section{The Scheduler (3)}

@@

\section{Scheduler PMC (1)}

@@

\section{Runloops (1)}

>>>>

\chapter{Acknowledgements}

I want to thank:
\begin{itemize}
\item{Andrew "whiteknight" Whitworth for laying the foundations for my work, being of tremendous help at all stages of the project and for fixing some of my bugs.}
\item{Nat "Chandon" Tuck for implementing the green thread basics on which my work is based on.}
\item{Christoph "cotto" Otto for bringing me into the Parrot project.}
\item{Brian "benabik" Gernhardt for lurking in the chat channel and at least trying to help with my many questions.}
\item{Markus "zimmski" Zimmerman for correcting many errors.}
\end{itemize}

\chapter{Abstract}

Parrot is a runtime system for dynamically typed programming languages. Despite several attempts in it's 10 years of history, it does not provide any support for multithreaded computation.

This thesis shows a way to implement threading support in Parrot using a hybrid approach using a combination of lightweight (``green'') threads and operating system threads. These lightweight threads are used as messages in a system where reading shared data is allowed but only the one owner thread may write to it. The implementation of this hybrid approach is described in detail and benchmarks are presented demonstrating the viability of this design.

\begin{german}
\chapter{Kurzfassung}

Parrot ist eine Laufzeitumgebung f{\"u}r dynamisch typisierte Programmiersprachen. Trotz mehrerer Anl{\"a}ufe in seiner zehnj{\"a}hrigen Geschichte, bietet Parrot keine Unterst{\"u}tzung f{\"u}r Multithreading.

Diese Arbeit zeigt einen Weg, um Threadingunterst{\"u}tzung in der Laufzeitumgebung zu implementieren mit Hilfe eines hybriden Systems aus leichtgewichtigen und Betriebssystem-Threads. Diese leichtgewichtigen Threads werden als Nachrichten in einem System benutzt, in dem das Lesen von gemeinsamen Daten erlaubt ist, aber nur der eine besitzende Thread darauf schreiben darf. Die Implementierung wird detailiert beschrieben und mit Benchmarks die Brauchbarkeit des Designs gezeigt.

@@

\chapter{Introduction and Motivation}

%
On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a flexible and widely used programming language but had started to show its age and suffered from early design decisions.

Am example of Perl:
\input{perlexample}

The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility \cite{Masak40451}. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions \cite{Developer}. Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot \cite{Perl}.

Parrot evolved from being just the interpreter for the new version of Perl to being a language independent VM providing features like garbage collection, exception handling and dynamic typing. At the time when the Parrot project started, the Java and .NET VM were widely used, but both targeted statically typed languages. Parrot thus filled a quickly growing niche.

The Perl 6 design process began with asking the Perl users what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multithreading support \cite{RFC1}. Perl 5 had two different implementations of thread support. In the first model, called \textit{5005threads}, data was shared by default and shared access to data had to be explicitly synchronized. This was similar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use \cite{ThreadManual}. Perl 5.6 introduced the newer model called \textit{ithreads}, mostly as a way to emulate \textit{fork} on Win32 platforms. Perl 5.8 exposed this Application Programming Interface (API) to the user of the programming language. In this new model, all data is copied to each thread and afterwards thread local. Data must be explicitly shared between threads.

In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

Being born at a time when Perl 6 still looked much more similar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's \textit{ithreads} model. Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like \textit{Software Transactional Memory} failed. For example Parrot has never supported running multiple threads and having garbage collection at the same time.

@@

In the year 2005 development of faster Central Processing Units (CPUs) shifted from increased speed of a single core to adding more cores. Modern processors contain up to 12 cores with even mobile phones having up to four. To utilize a modern CPU's power, code needs to be run in parallel. In UNIX (and thus Perl) tradition, this is accomplished using multiple processes being a good solution for many use cases. For many others like auto threading of hyper operators in Perl 6, the cost of process setup and communication would be prohibitively high except for very large data sets.

@@

Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption and synchronization entirely to the user. A high-level language like Perl 6 on the other hand provides complex and compound data types, handles garbage collection and a very dynamic object system. Even seemingly simple things like a method call can become very complex. In a statically typed programming language the definition of a class is immutable. Thus, calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it. Calling the same method again can then even omit the first two steps since their results cannot change.

In a dynamic language, the object may change its class at runtime. The inheritance hierarchy of the class may be changed by adding or removing parent classes. Methods may be added to or removed from classes (or objects) at runtime and even the way how to find a method of a class may change. So a simple method call results in the following steps:
%
\begin{itemize}
\item determining the class of the object,
\item determining the method resolution method of the class,
\item finding the actual method to call,
\item calling the method.
\end{itemize}
%
These steps have to be repeated for every method call, because the results may change any time. In a threaded environment, a thread running in parallel may change the underlying data and meta data in between those sequences and even between those steps. As a consequence, this meta data has to be protected from corruption introducing the need for locks in a performance critical area.

Many interpreters for dynamic languages like Python \cite{LearningPython} or Ruby \cite{TheRubyProgrammingLanguage} handle this problem by using a global interpreter lock to effectively serialize all operations. This is a proven and reliable way but leaves much of the hardware's potential unused.

@@

During the years of back and forth and failed attempts of adding threading support to Parrot, the Perl 6 specification evolved to a point where the largest parts of the language were covered and its features were implemented in the compilers. The lack of concurrency primitives in Parrot however prevents any progress in the area of concurrency support.

Before the work on this thesis started, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.

This thesis suggests a new approach based on a hybrid threading system. So called \textit{green threads} are used to simplify the implementation of a nearly lock free multithreading implementation. This approach is based on a design by Andrew Whitworth and Nat Tuck \cite{WKnightThreading}. The goal of this thesis is to demonstrate the advantages of this model and produce a working implementation which can be used to investigate the performance characteristics of a hybrid threading system.

\chapter{Concurrency in other Programming Platforms}

This chapter is about programming platforms. A platform is seen as a combination of a programming language and a runtime. E.g. for the Python programming language there are multiple runtimes with different implementations of threading support.

@@

In Java, the user is responsible for preventing concurrency issues. The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect the consistency of the provided data structures. The class library provides the user with high-level data structures explicitly designed for multithreaded scenarios.

Java version 1.1 used \textit{green threads} to support multithreaded execution of Java programs. Green threads are threads simulated by the virtual machine (VM) but unable to use more than one CPU core for processing. Details are described in chapter~\ref{cha:green_threads}. Version 1.2 introduced native Operating System (OS) threading support which since has become the standard way to do multithreading in Java \cite{JavaThreadManual}.

@@

The CPython implementation of the Python runtime uses a \textit{Global Interpreter Lock} (GIL) to protect its internal consistency \cite{PythonThreadingManual}. This is a single lock taken whenever the interpreter executes Python bytecode. Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitly type safe. The drawback is that Python code cannot benefit from having multiple CPU cores available. However I/O operations and calls to external libraries are executed without holding the GIL, so in applications with multiple I/O bounded threads, there may still be a performance benefit from using multithreading.

To run Python code in parallel, multiple processes have to be used. The \textit{multiprocessing} module provides support for spawning processes exposed through an API similar to the \textit{threading} module \cite{PythonMultiProcessingManual}. Since processes may not directly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.

\chapter{Parrot}

Parrot consists of the VM (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the \textit{Parrot Compiler Toolkit}). This thesis concentrates on the VM itself. The interpreter is written in C. Example code and test cases are written in \textit{Parrot Intermediate Representation} (PIR) a high-level assembly language abstracting register allocations and function calling conventions.

Contrary to other widely used VMs like the JVM or the \textit{Common Language Runtime} (CLR) which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based. A stack based VM usually pops the operands for an operation from the top of a stack and pushes the result back. Thus the operands are chosen implicitly by ordering of the operations allowing the opcodes to be without operands. In a register based VM on the other hand, each operation has to specify the operands explicitly. Compilers for stack machines are simpler because they do not have to care about register allocation and code is independent of prior or subsequent code \cite{VMShowdown}. The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just-in-time (JIT) compilation and improved performance of nested function and method calls.

The current design of Parrot uses four sets of registers where each one has an unlimited number of registers. The four sets are correspondingly to Parrot's types:
%
\begin{itemize}
\item integer,
\item floating point,
\item string and
\item polymorphic container (PMC).
\end{itemize}

The first three register types are self explanatory. String in Parrot is a low-level type with the interpreter handling all memory allocation issues and Unicode encoding. String values are immutable.

\section{PolyMorphic Containers (PMCs)}
\label{sec:PMCs}

PMCs are containers for all high level types such as objects, arrays, hash tables or code. Thus, they are similar to Python's \textit{PythonObject} types. PMCs are defined by C structs and are garbage collected. Their definition looks like:
\begin{CCode}
struct PMC {
    Parrot_UInt    flags;
    VTABLE         *vtable;             /* Pointer to vtable. */
    DPOINTER       *data;               /* Pointer to attribute structure. */
    PMC            *_metadata;          /* Pointer to metadata PMC. */
};
\end{CCode}
In short, they contain a pointer to a type specific data structure and some meta data. The \textit{vtable} is a virtual method table, where the type's behavior is defined. It contains a list of function pointers forming Parrot's unified type interface. This interface is a union of numeric, string, array, hash and object like behaviors. For example the  \textit{get\_integer} function returns an integer value for the data type. For a simple \textit{int} it is the value. For an array it may be the number of elements. The \textit{find\_method} function makes user defined methods of objects possible.

In this way, a language implementer can define the basic types of the language and available operations on them. For each \textit{vtable} entry, there is a corresponding opcode in Parrot's bytecode. Thus an \textit{inc \$P0} instruction calls the \textit{increment} vtable function of the PMC pointed to by the \textit{\$P0} register.

PMCs are implemented in \textit{pmc} files with method bodies written in C. These files are preprocessed to plain C before compilation.

\section{\textit{ParrotInterpreter}}

The Parrot interpreter is represented by a C struct called \textit{parrot\_interp\_t}. This structure contains pointers to the garbage collector's runtime data, the loaded types, vtables, the runloop, the current continuation and other global data. A pointer to this structure is passed to almost every function as the first parameter.

There is also the \textit{ParrotInterpreter} PMC which exposes parts of the \textit{parrot\_interp\_t} struct to the user.

@@

Control flow in Parrot is modeled using continuation passing. A continuation is a data structure containing the state of a program at a given point in its execution. It contains all information necessary to continue a program in a certain state, e.g. a call stack, the instruction pointer and contents of local variables. Calling a continuation means restoring the state encapsulated in the continuation.

When a function is called, instead of pushing a return address on the stack, the function is given a return continuation as part of its parameters. Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them. Continuation passing makes things like tail call optimization\footnote{When in the last statement of a function another function is called and its value returned unchanged, the outer function can just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function.} simple and as described in chapter~\ref{cha:green_threads} is a important part of implementing green threads.

@@

A runloop is the inner most loop executing bytecode in Parrot. It is also a data structure containing some information needed to support exception handling. When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started. Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code calling a previously defined exception handler in user code.

@@

Before entering a runloop, the \textit{setjmp} C function is used to save the current stack position and register contents in a data structure stored in the runloop meta data. This is effectively creating something resembling a continuation at the C level. When an exception is created within the interpreter, the runloop stack is searched for the runloop containing a suitable exception handler. \textit{longjmp} is then used to unwind the call stack up to the point where \textit{setjmp} was called and the call environment is restored. While having great similarities with continuations, this mechanism is more limited. It only allows to jump back to a point ``higher'' in the call stack.

\section{Garbage Collector (GC)}

Parrot supports different Garbage Collector (GC) implementations which can be selected at interpreter startup. Currently, there are four implementations of different algorithms:
\begin{itemize}
\item Inf: a GC for debugging purposes never collecting any garbage.
\item MS: a basic mark and sweep implementation.
\item MS2: a non-recursive mark and sweep implementation.
\item GMS: a generational, non-compacting, mark and sweep GC.
\end{itemize}
GMS is the default.

A mark and sweep garbage collector operates in two phases: In the mark phase, starting from a known root set of objects, the GC follows pointers in the object graph, marking each encountered object as alive. In the sweep phase, all objects not being alive are destroyed and all the live flags are reset.

In a process with many objects, having to traverse the whole graph may take a considerable amount of time. To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object is alive, the lower the chances are that it will become unused. So, the objects are partitioned into different generations. The youngest generation will always be traversed while the older generations will be handled much less frequently or even never more at all.

\section{Historical development}

Much of Parrot's previous threading related code has been removed to clean up the code and improve performance. Since the existing threading support was known to be unreliable and seriously flawed, this was no trade off. The final parts were removed by the merging of the \textit{kill\_threads} branch on September, 21st 2011.

In 2010, Nat Tuck began working on a \textit{green\_threads} branch during his Google Summer of Code internship. The feature got prototyped using pure PIR and then implemented in Parrot's core. He got it to work in simple cases and started to work on operating system thread support but the internship ended before the code was ready to be merged into the master branch. The code lay dormant until the work on this thesis started in 2011.

\chapter{Green threads}
\label{cha:green_threads}

@@

OS threads are like multiple processes running at the same time but share their memory. The operating system is responsible for giving each of these threads CPU time to run. OS threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.

@@

"Green threads" or "lightweight threads" are threads managed by the VM instead of by the operating system. For better distinction green threads are henceforth referred to as "tasks". These tasks only run pseudo parallel. The VM contains a scheduler and support for preempting running tasks.
Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
This is an implementation of the many-to-one threading model very similar to what an operating system running on a system with a single CPU core does.
%
Advantages of green threads are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item Green threads are very light weight with low memory overhead and close to zero creation time.
\item They do not depend on OS threading support
\item The interpreter controls the point of preemption of a green thread.
\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
\item Garbage collection can be implemented like in a single threaded process.
\item Critical sections can be protected by disabling the scheduler until the code is clear of the section.
\end{itemize}
%
Disadvantages include:
%
\begin{itemize}
\item They do not allow more than one CPU core to be used for computations.
\item Blocking calls like I/O block the whole interpreter including other green threads.
\item The interpreter has to contain logic and timers to control green threads.
\end{itemize}

@@

\chapter{Design of Hybrid Threads}
\label{cha:design}

@@

In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread. Thus the words \textit{thread} and \textit{interpreter} are used interchangeably as there is a 1:1 relationship between them. When a user starts a new task, the scheduler first looks for an idle thread. If one can be found, the task is scheduled on the thread's interpreter. If none can be found, a new thread with a new interpreter is started. Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system. If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters. This is effectively an implementation of the N:M threading model.

@@

As described in the introduction, cross thread writes to shared data may endanger the internal consistency of the interpreter. Traditionally the response to this problem is the use of locks of varying granularity. Fine grained locking allows code to run well parallelized but taking and releasing locks also costs performance. It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate. Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}. While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.

@@

The solution for these problems implemented for this thesis is to side step them altogether by disallowing write access to shared data. The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started. The interpreter would then create proxy objects for these variables which the task can use to access the data. These proxies contain references to the original objects. They use these references to forward all reading \textit{vtable} functions to the originals. Write access on the other hand would lead to a runtime error.

@@

For threads to be able to communicate with their creators and other threads, they still need to write to shared data. This is where green threads come into play. Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable. This task would be scheduled on the interpreter owning this variable. To reduce latency, the task is flagged to run immediately. The data owning interpreter would preempt the currently running task and process the new write task. Put another way, the data owning interpreter is told what to write to its variables, so other threads don't have to.

@@

\chapter{Implementation of Hybrid Threads}

\section{The Scheduler}

@@

\section{Scheduler PMC}

@@

\section{Runloops}
